Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data

### 2021-07-27 paper/paper.Rmd
Ready to submit.

---
title: |
  Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data
type: Article
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
  - name: Stuart Lee
    affil: b, c
    email: stuart.lee1@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
  - num: c
    address: |
      Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  In high-dimensional data analysis the curse of dimensionality reasons that points tend to be far away from the center of the distribution and on the edge of high-dimensional space. Contrary to this, is that projected data tends to clump at the center. This gives a sense that any structure near the center of the projection is obscured, whether this is true or not. A geometric transformation to reverse the curse, is defined in this paper, which uses radial transformations on the projected data. It is integrated seamlessly into the grand tour algorithm, and we have called it a burning sage tour, to indicate that it reverses the curse. The work is implemented into the tourr package in R. Several case studies are included that show how the sage visualizations enhance exploratory clustering and classification problems.
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics; data science; machine learning
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
keep_tex: true
---

@@

Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution. This affects visualizations made on low-dimensional projections. 
The effect is described by @diaconis1984, \textcolor{red}{showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.}
This has motivated the development of indexes for projection pursuit which search for departure from normality. It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976;@10.1093/biomet/asp084]: all observations can collapse into a single point.  These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

@@

The aim of the algorithm is to redistribute the projected volume such that equal relative areas on the disk, as given by $v_{2} (r; p=2, R)= v_p(r; p=2, R) = (r/R)^2$, contain equal relative projected volume, given by $v_{2} (r; p, R)$. This is achieved through a transformation of the projected radius that can be defined for any $r\in[0,R]$, and is applied to the projected data points in the plane, $y = (y_1, y_2)$. We work with polar coordinates and represent the data points as $y = (r_y, \theta_y)$. The angular component $\theta_y$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed. The radial component $r_y$ is transformed in two steps.

\textcolor{red}{We transform $r_y$ in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to } 
transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk. This inverse is defined via $v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y$ and thus $v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}$.

@@

\textcolor{red}{At this point a few characteristics of the transformation should be noted. First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph. In addition, since only the radial component is transformed, angles between the points will not be altered. (This can also be shown explicitly, from mathematical reformulation of the angles between two points after the transformation.) Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$). For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space. Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}

@@

\textcolor{red}{Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in densitiy near the center. When studying the separation between groups it is important to keep in mind that the effect is the same within and between clusters. Thus, if points within a cluster remain grouped while separation between clusters becomes more apparant, this can be interpreted as a true finding. Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.}

\textcolor{red}{We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane. We could also think about this is in terms of the distribution function: a uniform distribution in high-dimensions should still look uniform in the projected view. Our approach is defining the transformation assuming that points are uniformly distributed in the hypersphere. This could be generalized to work with the empirical radial cumulative distribution function instead.}

>>>>

---
title: |
  Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data
type: Article
author:
  - name: Ursula Laa
    affil: a, b, c
    email: ursula.laa@boku.ac.at
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
  - name: Stuart Lee
    affil: b, d
    email: stuart.lee1@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
  - num: c
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences
  - num: d
    address: |
      Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  In high-dimensional data analysis the curse of dimensionality reasons that points tend to be far away from the center of the distribution and on the edge of high-dimensional space. Contrary to this, is that projected data tends to clump at the center. This gives a sense that any structure near the center of the projection is obscured, whether this is true or not. A geometric transformation to reverse the curse, is defined in this paper, which uses radial transformations on the projected data. It is integrated seamlessly into the grand tour algorithm, and we have called it a burning sage tour, to indicate that it reverses the curse. The work is implemented into the tourr package in R. Several case studies are included that show how the sage visualizations enhance exploratory clustering and classification problems.
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics; data science; machine learning
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
keep_tex: true
---

@@

Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution. This affects visualizations made on low-dimensional projections. 
The effect is described by @diaconis1984, showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.
This has motivated the development of indexes for projection pursuit which search for departure from normality. It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976;@10.1093/biomet/asp084]: all observations can collapse into a single point.  These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

@@

The aim of the algorithm is to redistribute the projected volume such that equal relative areas on the disk, as given by $v_{2} (r; p=2, R)= v_p(r; p=2, R) = (r/R)^2$, contain equal relative projected volume, given by $v_{2} (r; p, R)$. This is achieved through a transformation of the projected radius that can be defined for any $r\in[0,R]$, and is applied to the projected data points in the plane, $y = (y_1, y_2)$. We work with polar coordinates and represent the data points as $y = (r_y, \theta_y)$. The angular component $\theta_y$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed.

The radial component $r_y$ is transformed in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to 
transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk. This inverse is defined via $v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y$ and thus $v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}$.

@@

At this point a few characteristics of the transformation should be noted. First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph. In addition, since only the radial component is transformed, angles between the points will not be altered. (This can also be shown explicitly, from mathematical reformulation of the angles between two points after the transformation.) Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$). For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space. Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.

@@

Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in density near the center. When studying the separation between groups it is important to keep in mind that the effect is the same within and between clusters. Thus, if points within a cluster remain grouped while separation between clusters becomes more apparent, this can be interpreted as a true finding. Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.

We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane. We could also think about this is in terms of the distribution function: a uniform distribution in high-dimensions should still look uniform in the projected view. Our approach is defining the transformation assuming that points are uniformly distributed in the hypersphere. This could be generalized to work with the empirical radial cumulative distribution function instead.

### 2021-07-27 paper/paper.tex
Ready to submit.

\author{\name{Ursula Laa$^{a, b}$, Dianne Cook$^{b}$, Stuart Lee$^{b,
c}$}
\affil{$^{a}$School of Physics and Astronomy, Monash
University; $^{b}$Department of Econometrics and Business Statistics,
Monash University; $^{c}$Molecular Medicine Division, Walter and Eliza
Hall Institute, Parkville, Australia}
}

\thanks{CONTACT Ursula
Laa. Email: \href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}, Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Stuart
Lee. Email: \href{mailto:stuart.lee1@monash.edu}{\nolinkurl{stuart.lee1@monash.edu}}}

@@

Paradoxically, the curse of dimensionality inverts for dimension
reduction, resulting in an excessive amount of observations near the
center of the distribution. This affects visualizations made on
low-dimensional projections. The effect is described by
\citet{diaconis1984},
\textcolor{red}{showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.}
This has motivated the development of indexes for projection pursuit
which search for departure from normality. It is also related to what is
called ``data piling'' in high-dimension low-sample size data
\citep{10.2307/27639976, 10.1093/biomet/asp084}: all observations can
collapse into a single point. These issues also persist with non-linear
dimension reduction techniques, and are often referred to as the
``crowding problem'', which methods like t-Distributed Stochastic
Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.

@@

The aim of the algorithm is to redistribute the projected volume such
that equal relative areas on the disk, as given by
\(v_{2} (r; p=2, R)= v_p(r; p=2, R) = (r/R)^2\), contain equal relative
projected volume, given by \(v_{2} (r; p, R)\). This is achieved through
a transformation of the projected radius that can be defined for any
\(r\in[0,R]\), and is applied to the projected data points in the plane,
\(y = (y_1, y_2)\). We work with polar coordinates and represent the
data points as \(y = (r_y, \theta_y)\). The angular component
\(\theta_y\) is uniform for this distribution, by the rotation
invariance of the sphere, and thus does not need to be transformed. The
radial component \(r_y\) is transformed in two steps.

\textcolor{red}{We transform $r_y$ in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to }
transform \(v_{2} (r_y; p, R)\) using the inverse of
\(v_{2} (r_y; 2, R)\), to go from a uniform distribution in radius to a
uniform distribution in area of the disk. This inverse is defined via
\(v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y\)
and thus \(v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}\).

@@

\textcolor{red}{At this point a few characteristics of the transformation should be noted. First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph. In addition, since only the radial component is transformed, angles between the points will not be altered. (This can also be shown explicitly, from mathematical reformulation of the angles between two points after the transformation.) Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$). For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space. Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}

>>>>

\author{\name{Ursula Laa$^{a, b, c}$, Dianne Cook$^{b}$, Stuart Lee$^{b,
d}$}
\affil{$^{a}$School of Physics and Astronomy, Monash
University; $^{b}$Department of Econometrics and Business Statistics,
Monash University; $^{c}$Institute of Statistics, University of Natural
Resources and Life Sciences; $^{d}$Molecular Medicine Division, Walter
and Eliza Hall Institute, Parkville, Australia}
}

\thanks{CONTACT Ursula
Laa. Email: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}, Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Stuart
Lee. Email: \href{mailto:stuart.lee1@monash.edu}{\nolinkurl{stuart.lee1@monash.edu}}}

@@

Paradoxically, the curse of dimensionality inverts for dimension
reduction, resulting in an excessive amount of observations near the
center of the distribution. This affects visualizations made on
low-dimensional projections. The effect is described by
\citet{diaconis1984}, showing that most linear projections onto
low-dimensional subspaces are approximately Gaussian, and that
observations concentrate near the center as the dimensionality of the
full space increases. This has motivated the development of indexes for
projection pursuit which search for departure from normality. It is also
related to what is called ``data piling'' in high-dimension low-sample
size data \citep{10.2307/27639976, 10.1093/biomet/asp084}: all
observations can collapse into a single point. These issues also persist
with non-linear dimension reduction techniques, and are often referred
to as the ``crowding problem'', which methods like t-Distributed
Stochastic Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.

@@

The aim of the algorithm is to redistribute the projected volume such
that equal relative areas on the disk, as given by
\(v_{2} (r; p=2, R)= v_p(r; p=2, R) = (r/R)^2\), contain equal relative
projected volume, given by \(v_{2} (r; p, R)\). This is achieved through
a transformation of the projected radius that can be defined for any
\(r\in[0,R]\), and is applied to the projected data points in the plane,
\(y = (y_1, y_2)\). We work with polar coordinates and represent the
data points as \(y = (r_y, \theta_y)\). The angular component
\(\theta_y\) is uniform for this distribution, by the rotation
invariance of the sphere, and thus does not need to be transformed.

The radial component \(r_y\) is transformed in two steps. First, we
replace it with \(v_{2} (r_y; p, R)\) to redistribute the volume
uniformly by the radius. The second step is then to transform
\(v_{2} (r_y; p, R)\) using the inverse of \(v_{2} (r_y; 2, R)\), to go
from a uniform distribution in radius to a uniform distribution in area
of the disk. This inverse is defined via
\(v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y\)
and thus \(v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}\).

@@

At this point a few characteristics of the transformation should be
noted. First, as already mentioned, it is approximately linear near the
center, and thus will not affect shapes in that part of the graph. In
addition, since only the radial component is transformed, angles between
the points will not be altered. (This can also be shown explicitly, from
mathematical reformulation of the angles between two points after the
transformation.) Note however that shapes will be affected further away
from the center (in particular towards the maximum range \(R\)). For
example, a linear relation might appear curved because of the non-linear
nature of the transformation in that region of the projected space.
Thus, while the interpretation is straightforward for points projected
in the central region, care must be taken with the interpretation of
features observed in the outer region of the projected distribution.
